### 2025-06-11 talk/cmo25.tex
add cmo25 talk

\begin{center}
  {\Large\bf An iterated block particle filter for inference on coupled dynamic systems}

% ABSTRACT
% We consider inference for a collection of partially observed, stochastic, interacting, nonlinear dynamic processes. Each process is called a unit, and our primary motivation arises in biological metapopulation systems where a unit is a spatially distinct sub-population. Block particle filters are an effective tool for simulation-based likelihood evaluation for these systems, which are strongly dependent through time on a single unit and relatively weakly coupled between units. Iterated filtering algorithms can facilitate likelihood maximization for simulation-based filters. We introduce a new iterated block particle filter algorithm applicable to parameters that are either unit-specific or shared between units. We demonstrate this algorithm to carry out inference on a coupled epidemiological model for spatiotemporal COVID-19 case report data in 373 cities.

@@

IMS Asia Pacific Rim Meeting\\
Jan 5, 2024

@@

Slides are at \url{https://ionides.github.io/talks/aprm24.pdf}

@@

Joint work with
Patricia Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual

@@

%  \vspace{2mm}

%  \item Bonus question: How do we know if our model is statistically adequate, or needs more work?

@@

      \url{https://ionides.github.io/531w22/}

@@

@@

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Li et al (2020) used iterated EnKF for inference.
  \item The time interval covers the initial China lockdown.
  \item We present re-analysis of this model and data, recently posted on arXiv.
%% see README for R code
%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
%   \item{} 0 cases reported for all cities during Jan 10--15.

@@

\begin{frame}
  %%{An SEAIR metapopulation model for COVID-19}
\begin{center}
%%%%% SEAIR diagram
\resizebox{11cm}{!}{
\begin{tikzpicture}[
  square/.style={rectangle, draw=black, minimum width=0.8cm, minimum height=0.8cm, rounded corners=.1cm, fill=blue!8},
  travel/.style={circle, draw=black, minimum width=0.85cm, minimum height=0.8cm, fill=green!8},
  report/.style={shape=regular polygon, regular polygon sides=8, draw, fill=red!8,minimum size=0.9cm,inner sep=0cm},
  bendy/.style={bend left=25},
  >/.style={shorten >=0.25mm}, % redefine arrow to stop short of node
  >/.tip={Stealth[length=1.5mm,width=1.5mm]} % redefine arrow style
]
\tikzset{>={}}; % this is needed to implement the arrow redefinition

\node (S) at (-0.5,0) [square] {S$_u$};
\node (E) at (2,0) [square] {E$_u$};
\node (A) at (4.5,0) [square] {A$_u$};
\node (I) at (4.5,-1.75) [square] {I$_u$};
\node (R) at (7,0) [square] {R$_u$};

\node (T1) at (-0.5,1.75) [travel] {T};
\node (T2) at (2,1.75) [travel] {T};
\node (T3) at (4.5,1.75) [travel] {T};

\node (Ca) at (4.5,-3.5) [report] {C$^a_u$};
\node (Cb) at (7,-3.5) [report] {C$^b_u$};
\node (C) at (9.5,-3.5) [report] {C$_u$};

\node (V1) at (2.5,-1.75) {};
\node (V2) at (3,-3.5) {};
\draw [->] (E) -- (Ca -| V2) -- (Ca);
\draw [->] (I -| V1) -- (I);

\draw [->] (S) -- (E);
\draw [->] (A) -- (R);

\draw [->] (Ca) -- (Cb);
\draw [->] (Cb) -- (C);

\draw [->] (I.east) -- (R);
\draw [->] (E) -- (A);

\draw [->, bendy] (S) to (T1);
\draw [->, bendy] (T1) to (S);

\draw [->, bendy] (E) to (T2);
\draw [->, bendy] (T2) to (E);

\draw [->, bendy] (A) to (T3);
\draw [->, bendy] (T3) to (A);

\end{tikzpicture}
}
\end{center}
\bi
\item 
  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
  \item 
An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
\item
  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
  \item 
Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
\ei

@@

\begin{frame}
 \includegraphics[width=12cm]{covid/filter_tests10.png}

@@

  \includegraphics[trim={0 0 0 10mm},clip,width=9cm]{IBPF_workflow.pdf}

@@

\nocite{asfaw23arxiv,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides22,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler23,li23}

\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{bib}
\end{frame}

>>>>

\begin{center}
  {\Large\bf Inference on spatiotemporal infectious disease dynamics}

%ABSTRACT: Epidemiological models must be consistent with observed data in order to generate reliable knowledge and evidence-based policy. Metapopulation systems, which consist of a network of connected sub-populations, pose technical challenges in statistical inference owing to nonlinear, stochastic interactions. Numerical difficulties encountered in conducting inference can obstruct the core scientific questions concerning the link between the mathematical models and the data. Recently, an algorithm has been proposed that enables computationally tractable likelihood-based inference for high-dimensional partially observed stochastic dynamic models of metapopulation systems. We use this algorithm to build a statistically principled data analysis workflow for metapopulation systems. Via a case study of COVID-19, we show how this workflow addresses the limitations of previous approaches. The COVID-19 pandemic provides a situation where mathematical models and their policy implications were widely visible, and we revisit an influential metapopulation model used to inform basic epidemiological understanding early in the pandemic. Our methods support self-critical data analysis, enabling us to identify and address model weaknesses, leading to a new model with substantially improved statistical fit and parameter identifiability. Our results suggest that the lockdown initiated on 23 January 2020 in China was more effective than previously thought.

@@

Workshop on {\it Mathematical and Statistical Challenges in \\
  Post-Pandemic Epidemiology and Public Health}\\
Casa Matem\'{a}tica Oaxaca\\
June 16, 2025

@@

Slides are at \url{https://ionides.github.io/talks/cmo25.pdf}

@@

Joint work with
Ning (Patricia) Ning, Jesse Wheeler, Kidus Asfaw, Jifan Li, Joonha Park, Aaron King, Mercedes Pascual

@@

\item Bonus question: How do we know if our model is statistically adequate, or needs more work? Hint: we need {\bf performance benchmarks}, i.e., comparison with non-mechanistic machine learning methods. If there is performance shortfall, we need diagnostics to figure out where and how.

@@

      \url{https://ionides.github.io/531w24/}

@@

\begin{frame}{\texttt{spatPomp}: an R package for SpatPOMP models}

  Asfaw et al ({\it J. Open Source Software}, 2024)
  \url{https://github.com/spatPomp-org}

  \bi
\item Permits specification of arbitrary non-linear non-Gaussian spatiotemporal POMP models.
\item Adds spatial structure to \texttt{pomp} \citep{king16}.
\item Monte Carlo algorithms designed to address the \myemph{curse of dimensionality}.
  \item Algorithms compile via C to assist performance.
\item Statistically efficient, likelihood-based inference despite methods being simulation-based.
  \ei

  Future goals: add GPU and autodiff support via \texttt{pypomp}, \url{https://github.com/pypomp}

@@

  \begin{itemize}
  \item Metapopulation data were used to infer the fraction of asymptomatic cases and their contagiousness
    (Li et al, {\it Science}, May 2020).
  \item SEIR (susceptible-exposed-infected-removed) model with asymptomatics, reporting delay, and coupling based on cell phone data.
  \item Li et al (2020) used iterated EnKF for inference.
  \item The time interval covers the initial China lockdown.
  \item We present re-analysis of this model and data (Li et al, {J. Roy. Soc. Interface}, 2024).
%% see README for R code
%   \item Total cases 801, Wuhan 454, Chongqing 27, Beijing 26.
%   \item{} 259 cities with 0 cases, 99 with 1--5 cases, 17 with $>5$ cases.
%   \item{} 0 cases reported for all cities during Jan 10--15.

@@

\begin{frame}
  %%{An SEAIR metapopulation model for COVID-19}
  \begin{center}
      \begin{center}
    \includegraphics[height=5cm]{seair.png}
    \end{center}

\end{center}
\bi
\item 
  Reportably infectious individuals, $\mathrm{I}_u$ for city $u$, are included in the delayed reporting compartment, $\mathrm{C^a_u}$.
  \item 
An individual arriving at $\mathrm{C}_u$ is a case report for city $u$.
\item
  Individuals in $\mathrm{A}_u$ are not reportable and transmit at a reduced rate.
  \item 
Travel occurs to and from $\mathrm{T}$, based on 2018 data from Tencent.
%\item Many models have been proposed. We want a framework to implement and evaluate various scientific hypotheses.
\ei

\begin{frame}{Log-likelihood values for six models}

  \resizebox{12cm}{!}{
\begin{tabular}{c|rr|l}
 & loglik & df & description
\\
\hline
$M_1$ &
  $-\infty$ & 10 &
  SEAIR model, parameter values and mobility of Li et al (2020)
\\
$M_2$ &  -14240.5 & 10 &
  Adjusted mobility and measurement in $M_1$
\\
$M_3$ &
 -11257.9  & 374 &
  Independent identically distributed negative binomial
\\
$M_4$ &
 -10825.3 & 375 &
  Autoregressive negative binomial
\\
$M_5$ &
  -9088.2&
  12 &
  Adding overdispersed dynamics to $M_2$ and refitting
\\
$M_6$ &
  -9116.5&
  10 &
  Latent and infectious periods unchanged by lockdown in $M_5$
\\

\hline
\end{tabular}
}

  \vspace{10mm}

Model comparisons by log-likelihood. The degrees of freedom (df) is the number of estimated parameters.

\begin{frame}{Data analysis progression}
  \bi
\item Some data errors meant that the original data was incompatible with the model.
\item Tracked down by assessing conditional log-likelihood for each unit and each time.
\item Even then, the likelihood was poor compared to non-mechanistic benchmarks. Including over-dispersion fixed that.
\item Some weak identifiability remained. extra assumptions can lead to plausible parameter estimates without much loss of model fit.
\item Mild outliers remain, e.g., Wuhan on day 18.  
  \ei

@@

\begin{frame}
 \includegraphics[width=10cm]{covid/filter_tests10.png}

@@

%  \includegraphics[trim={0 0 0 10mm},clip,width=9cm]{IBPF_workflow.pdf}
  \includegraphics[width=11cm]{ibpf.png}

@@

\nocite{asfaw24,bjornstad01,breto19,grenfell04,he10,katzfuss19,ionides21,ionides24,lee20,li20,ng02,ning23-ibpf,park20,rebeschini15,wheeler24,li24}

\nocite{asfaw24tutorial}
\begin{frame}[allowframebreaks]
\frametitle{References}
\bibliographystyle{apalike}
\bibliography{bib}
\end{frame}
